Whose Truth is it Anyway? An Experiment on Annotation Bias

### 2023-06-22 report/draft.Rmd
update analyses in report
Last additions
Some copyediting, adding a sentence on ChatGPT and two sentences on strict/lenient stances

---
output: 
  stevetemplates::article:
    fig_caption: true
    extra_dependencies: ["float"]
bibliography: references.bib
urlcolor: black
linkcolor: black
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage[para,online,flushleft]{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{hyperref}
   - \usepackage{array}   
   - \usepackage{caption}
   - \usepackage{graphicx}
   - \usepackage{epstopdf}
   - \usepackage{siunitx}
   - \usepackage{hhline}
   - \usepackage{calc}
   - \usepackage{tabularx}
   - \usepackage{fontawesome}
   - \usepackage{amsthm}
   - \newtheorem{hypo}{Hypothesis}
biblio-style: apsr
title: "Whose Truth is it Anyway? An Experiment on Annotation Bias"
author:
- name: Mariken A.C.G. van der Velden*
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Myrthe Reuver
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Wouter van Atteveldt
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Antse Fokkens
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Felicia Loecherbach
- affiliation: Center for Social Media & Politics, New York University
- name: Kasper Welbers
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
thanks: "* = Corresponding author, Replication files are available on the author's Github account (https://github.com/MarikenvdVelden/bias-experiment); Author contributions: a) designed the study: MACGvdV, WvA, AF, FL, MR, & KW; b) conducted the study: MACGvdV, FL & MR; c) data cleaning & analysis: MACGvdV; d) writing of the paper: MACGvdV"
anonymous: FALSE
abstract: "Information is key to inform the behavior of citizens, and thereby for the social scientists studying them. The democratization of data has lead to numerous possibilities to gather and analyze textual data. These enourmous amounts of data are typically handled by machine learning techniques to classify into meaningful variables. The performance of these models is evaluated based on a gold standard, created by human annotators. Having a high level of agreement between these annotators is key, but some suggest personal characteristics of annotators, like political ideology or knowledge, interfere. We show in two pre-registered experiments that XXX. Thereby [contribution]."
keywords: "Experiment, Annotation Bias, Ideology, Measurinig Political Position, Text-as-Data, Political Knowledge"
geometry: margin=1in
mainfont: cochineal
fontsize: 11pt
params:
  anonymous: ""
doublespacing: TRUE
endnote: no
pandocparas: TRUE
sansitup: FALSE
---

@@

Information is key to inform the behavior of citizens, and thereby for the social scientists studying them. Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing]. The democratization of data and advent of computational social science has paved the way for new possibilities to gather and analyze textual data [for a recent overview, see @vanatteveldt2022computational]. In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see @bender2016linguistic; @wei2023overview]. 
To determine the validity of such large scale, computational analyses, we rely on "gold standard" data, created by human annotators. 
If follows that the validity of these analyses hinges on the quality of these golden standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

Most data collection efforts to create gold standards assume that there is only one correct interpretation for every input example, and that disagreement between the annotators is something that needs to be dealt with at all costs [@aroyo2015truth]. 
A recent study [@van2021validity] demonstrates that using crowd-coding platforms is a good way to collect such golden standards without too much disagreement. These platforms have also been used for experiments, and the quality of the respondents has been a center of attention [@coppock2019generalizing; @clifford2015samples; @huff2015these; @berinsky2014separating]. While it has been argued that this data is of similar quality to data from a random sample of the population [@coppock2019generalizing], others have demonstrated that these online platforms are populated by people that are "unlike" the general public, being younger and holding more liberal values [@clifford2015samples; @huff2015these]. If the latter is true, this might cause a problem for the coding of political texts: @ennser2018impact report that coders of political texts incorporate prior beliefs about parties' issue stances into their coding decisions. The authors find that party labels cue coders to a stance. For example, coders are more likely to report a left-wing party to be pro-immigration and a populist right-wing party to be against based on the exact same sentence. *Is this actually bias or a diversity of view points? And how big of a problem does this bias/diversity of view points generate for scholars relying on golden standard data?*

There is a long history of text annotation in studies analyzing political text. While this yields lots of experiences as to how to train coders so that we get reliable hand-coded data, the procedure is expensive, protracted, and sometimes does not even get us the quality of data we need [@weber18]. While the crowd offers a solution to some of these issues [@van2021validity], the main underlying assumption that there is one correct interpretation for every input example remains untouched. At the same time, trends of polarization have shown that people do interpret information according to their ideological position. *Does this mean that some are right and others are wrong? Or is there an ideological difference in the ground truth?* These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced. Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations [for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd, @ying2022topics]. However, work in computational linguistics has shown that increased annotation instructions do not increase quality [@parmar2022don]. This leaves us between a rock and a hard place. Is there a potential bias in annotators that we should account for?

In this paper, we build upon the NLP literature on disagreement -- or bias -- in annotation [e.g., see @shen2021sounds; @geva2019we; @sommerauer2020penguin; @plank2014learning] and so-called perspectivism [@basile2021toward; @havens2022beyond] -- i.e. the adoption of methods that integrate the opinions and perspectives of the human subjects involved in the knowledge representation step of the machine learning processes [@basile2021toward]. This literature puts forward that disagreement can occur because of differences in ideological position or political knowledge [@shen2021sounds; @alkiek2022classification; @joseph2021mis]. This allows us to test the extent to which disagreement takes place, for what type of stances, as well as gives us directions on how to deal with the diversity in conceptions and the political heterogeneity that nowadays potentially occurs in our sample of annotators. To do so, we have fielded two high-powered pre-registered experiments (see [here](link) and [here](link)) in the Netherlands -- a low-level polarized country -- and in the U.S. -- a high-level polarized country testing the effect of ideological distance between the annotator and the political actor in the text (H1), the effect of overinterpretation based on political knowledge or ideological engagement (H2), and a offered solution of masking the political actor to mitigate the effects of ideology and knowledge (H3).[^1] In the experiments, we vary the level of specification with which a political actor takes a stance -- a declarative sentence versus a sentence where with some knowledge on politics, the stance might be inferred -- as well as whether the political actor is shown or masked with putting `[ACTOR]` instead of the political party. We do this for four different political issues: *Environment*, *Immigration*, *Tax Policy*, and *EU* (for the Dutch case) or *Foreign Policy* (for the American case). This country selection does not only allow us to showcase the scope conditions of disagreement due to different levels of political heterogeneity, but also differentiates between languages. English is not only the most dominant language for computational text analysis in the social science [@baden2022three; Dolinsky et al. 2023], crowd-coders do not need to be native speakers, given the dominance of English in our daily lives. This is different for Dutch, it is a language spoken by a smaller community, typically native speakers, yet still an often-enough researched case in computational text analysis in the social science [@baden2022three; Dolinsky et al. 2023].

@@

Our results demonstrate that overall sentence where with some knowledge on politics the stance might be inferred are really difficult for the crowd to annotate -- people overinterpret the position using their own knowledge of the world. This is problematic as these sentence are very common in political text -- like legislative debates or speeches -- as well as media reports. Moreover, our results also demonstrate that for these disagreements in the crowd to occur, the level of polarization needs to be high. We do find support for our hypotheses in the American context, but not in the Dutch context -- except for the situation where masking overcomes political knowledge, we do find support for that in the Dutch case, but not in the American one. Our findings thus underline the importance of taking disagreement seriously for the creation of gold standard text -- the bread-and-butter of all machine learning endaveours. We should look beyond the majority vote and modeling it in the data, because if an algorithm is trained on biased data from disagreeing annotators, it will reproduce and often exacerbate that bias when it is applied to new data [e.g., see @prost2019debiasing]. To be able to model these characteristic of annotators, we should survey the characteristics of annotators when using the crowd (see Webb-Williams et al. 2023 for a similar argument, yet different annotator characteristics).

@@

Generating large data-sets has become one of the main drivers of progress in natural language understanding. In studies of political communication, the most familiar annotation tasks involve identifying basic concepts. This often involve noting the topic of the text, the position of the actor or the tone of the text. A recent study [@van2021validity] demonstrates that using crowd-coding platforms is a good way to collect such large data-sets. However, having only a few workers annotate the majority of text of interest has raised concerns about data diversity and models' ability to generalize beyond the crowd-workers: In a series of experiments, @geva2019we show that often models do not generalize well to annotations from annotators that did not contribute to the training set, suggesting that annotator bias should be monitored during data-set creation. One such potential bias, especially in times of increasing polarization [@iyengar2019origins; @gidron2019toward; @boxell2022cross], is based on ideological position of the annotator. Given that annotators on crows-coding platforms tend to be younger and hold more liberal values than the general public [@clifford2015samples; @huff2015these], this could potentially hamper the data diversity and generalizability of the model. An additional reason to monitor the annotators ideological position as a potential source of annotator bias is that a recent study in NLP showed that experiential factors influence the consistency of how political ideologies are perceived [@shen2021sounds]. Their finding challenges the "ground-truth" assumption we as researcher make that a position for example is either left-leaning or right-wing leaning. People with different ideological backgrounds might experience that position differently. 
This challenges our way of data collection.
We are interested in the effect of e.g. elite communication. To study this, we allow for heterogeneous treatment effects in experimental work. This indicates that we often do not assume that the treatment, often using text, has the same effect for different partisans. Yet, at the same time, we forget or ignore that knowledge when creating large data-sets for our machine learning models.

@@

Another aspect to consider is that disagreement can be informative for the concept under measure - sometimes agreement can be used to validate hypotheses about how universal the perceptions of such concepts are [@sommerauer2020penguin]. 
Additional doubts on smoothing out disagreement in annotation have focussed on the lack of diversity when only annotating with one label or annotator, leading to a homogeneity especially in subjective and social tasks [@geva2019we] such as hatespeech detection or political affiliation classification.
The question then is: _Whose perspective is being recorded in these datasets, and then later in the models trained on these datasets?_ 
Framing arbitrary representations in data as \"bias\" misses the political character of datasets: There is no neutral data and no apolitical standpoint from where we can call out bias. Datasets are always \"a worldview\" [26] and, as such, data always remains biased.\" (Miceli et al., 2022, p. 5).
This is key to social scientist in general, and those studying political text in particular, since several tasks of interest are intrinsically societal, with answers that differ based on the make-up of the worldview of annotators. 
The answer of the annotators in turn influence how machine learning examples classify new models. 
For instance, hate speech detection and abuse detection is one NLP task where the race and gender of annotators influences annotators and model performance[@gordon2022, @Larimore, @waseem2016]. 
Language is inherently connected to society and culture: @shenDarling analyze sentiment analysis, and find that human annotators lead to certain perspectives on sentiment being recorded, and that notably that African American English dialects are often misunderstood by such models.

Most recently, new annotation paradigms have gone one step further by asking whether we  modelling the task, or the annotator [@geva2019we]?
@pavlick2019 find that for the logical coherence task Natural Language Inference, annotators have several valid interpretations that are not reflected in one ground truth label. They call for new training paradigms that can reflect "the full range of possible human inferences"[@pavlick2019, p. 688]. Recent approaches in NLP have sought to explicitly incorporate disagreement and diversity in training data annotations. [@röttger2021] introduces the idea of an explicitly subjective annotation paradigm existing in addition to one focussed on one label and "ground truth". Such a subjective annotation paradigm can be used for purposes where the goal is finding diverse perspectives on the task or concepts, and for models to model more accurately how humans interpret a task or text. Additionally, "perspectivism" [@basile2021toward] is a paradigm and research agenda where different perspectives are explicitly incorporated in the training data, and used by models to provide more human-like classifications. Another paradigm is "jury learning" [@gordon2022], in which machine learning models do not learn to replicate one specific ground truth, but are trained with different annotator juries to reflect the judgement of different populations. In both approaches, demographic and other individual aspects of the annotator are explicitly mentioned and highlighted as having an influence on classification performance, but is used as an asset rather than as an error.

@troiano2021[p.2] also note how for a complex concept such as the "emotion" of a text, the annotator can make several assumptions during the annotation process on what is wanted, that are all valid and may or may not be useful in different contexts: "It is possible to assess one\'s own emotion after reading the text, to reconstruct the affective state of the writers who produced it, to guess the reaction that they intended to elicit in the readers, and so on.\" @shen2021sounds find that political ideology is not an inherent concept in many texts, but rather dependent on who is asked to annotate and their perceptions and background. Extralinguistic factors, such as annotator's own political ideology and also knowledge, influenced annotation and in turn model performance. @thornjakobsen2022 specifically analyze a task related to stance detection, argumentative sentence detection, on these extralinguistic factors, and find an effect of gender and political leaning on annotations and also model performance. However, to our knowledge this phenomenon has not yet been tested in a controlled experimental setting with manipulations in the texts.

@@

> **Ideological Bias hypothesis** (*H1b*): The effect of H1a is stronger for sentences in which the party's position can be infered (i.e. underspecified sentences).

As noted by @plank2014, some linguistic examples are ambiguous, and open to multiple interpretations even for seemingly objective tasks such as part of speech tagging, where annotators have to distinguish parts of speech such as nouns and verbs. A complex concept such as political ideology is much more likely to lead to multiple interpretations. We call such sentences with more possible interpretations and less explicit standpoints "underspecified". A lack of explicitness in the annotated text is one of the main causes of the disagreements in earlier literature. @thornjakobsen2022 deduce that annotator bias comes from a process known as the affect heuristic [@slovic2007affect ]: making a decision based on the emotional response related to your own personal attitude towards the discussed topic, especially when the text is relatively ambiguous.
To test whether people with different ideological backgrounds as well as their political knowledge might experience that position differently, challenging our ground-truth assumption in data annotation, we propose the following hypotheses:

@@

_If these biases exists, how can we alleviate them?_
There have been several previous approaches to solve biased annotations, especially where it concerns political or societal aspects. @geva2019we introduce an approach where training set annotators are separated from annotators annotating datasets that evaluate the models, to ensure the evaluation is not simply accurate at replicating the original annotators, but can generalize to new annotators' judgements. However, these approaches are not aimed at reducing biases during the training set annotation procedure. Other approaches are aimed at leveraging multiple perspectives - but this is not useful when one wants one label to learn from.
For the Austrian National Elections, @ennser2018impact already demonstrated that showing party labels impact annotators' assessment of the party position. So, one solution is masking:

@@

We have conducted the survey experiments in the Netherlands in May 2022 and in the United States in January 2023. 
Both samples, recruited through [KiesKompas](https://www.kieskompas.nl/en/) and [Prolific](https://www.prolific.co/) for respectively the Dutch and American case, consists of 3,000 participants (based on the power analysis presented in our [online compendium](LINK)) of 18 years and older. 
Both survey companies works with non-random opt-in respondents. 
Therefore, we measured many demographic background variables, and balance checks have been conducted to demonstrate whether certain categories are over represented in a certain experimental group. 
Our study has been approved by the [Research Ethics Review Committee](https://fsw.vu.nl/nl/onderzoek/research-ethics-review/index.aspx) of the *Vrije Universiteit Amsterdam* (see the approval [here](https://github.com/MarikenvdVelden/bias-experiment/blob/master/docs/2022-3-30-59.pdf)). To ensure good quality of our data, one attention check (discussed in more detail in [Section 3.3](#attention-checks)) is included [@berinsky2014separating].

@@

***Moderating Covariates.*** *Ideological position* is measured using an 11-point scale ranging from left (`0`) to right (`10`). *Political knowledge* is measured with six items from the Dutch Parliamentary Election Studies for the Dutch sample, and the three items from the American National Election Studies.[^]:The questionaire can be found in OA **XX**.

***Controle Variables.*** In our analysis, we control for demographic information (gender, age, education, income, religion, job) as well as political background variables (trust in politics, ideological position on economic left-right scale and cultural progressive-conservative scale, and evaluations and prospects of the economy). Tables A.5 till A.17 in the OA demonstrate the descriptive information per country.

>>>>

---
output: 
  stevetemplates::article:
    fig_caption: true
    extra_dependencies: ["float"]
bibliography: references.bib
urlcolor: black
linkcolor: black
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage[para,online,flushleft]{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{hyperref}
   - \usepackage{array}   
   - \usepackage{caption}
   - \usepackage{graphicx}
   - \usepackage{epstopdf}
   - \usepackage{siunitx}
   - \usepackage{hhline}
   - \usepackage{calc}
   - \usepackage{tabularx}
   - \usepackage{fontawesome}
   - \usepackage{amsthm}
   - \newtheorem{hypo}{Hypothesis}
biblio-style: apsr
title: "Whose Truth is it Anyway? An Experiment on Annotation Bias"
author:
- name: Mariken A.C.G. van der Velden*
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Myrthe Reuver
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Wouter van Atteveldt
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Antse Fokkens
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Felicia Loecherbach
- affiliation: Center for Social Media & Politics, New York University
- name: Kasper Welbers
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
thanks: "* = Corresponding author, Replication files are available on the author's Github account (https://github.com/MarikenvdVelden/bias-experiment); Author contributions: a) designed the study: MACGvdV, WvA, AF, FL, MR, & KW; b) conducted the study: MACGvdV, FL & MR; c) data cleaning & analysis: MACGvdV; d) writing of the paper: MACGvdV"
anonymous: FALSE
abstract: "Information is key to inform the behavior of citizens, and thereby for the social scientists studying them. The democratization of data has lead to numerous possibilities to gather and analyze textual data. These enourmous amounts of data are typically handled by machine learning techniques to classify into meaningful variables. The performance of these models is evaluated based on a gold standard, created by human annotators. Having a high level of agreement between these annotators is key, but some suggest personal characteristics of annotators, like political ideology or knowledge, interfere. We show in two pre-registered experiments that XXX. Thereby [contribution]."
keywords: "Experiment, Annotation Bias, Ideology, Measurinig Political Position, Text-as-Data, Political Knowledge"
geometry: margin=1in
mainfont: cochineal
fontsize: 11pt
params:
  anonymous: ""
doublespacing: TRUE
endnote: no
pandocparas: TRUE
sansitup: FALSE
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Information is key to inform the behavior of citizens, and thereby for the social scientists studying them.
Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing].
The democratization of data and advent of computational social science has paved the way for new possibilities to gather and analyze textual data [for a recent overview, see @vanatteveldt2022computational].
In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see @bender2016linguistic; @wei2023overview].
To determine the validity of such large scale, computational analyses, we rely on "gold standard" data, created by human annotators.
It follows that the validity of these analyses hinges on the quality of these golden standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

Most data collection efforts to create gold standards assume that there is only one correct interpretation for every input example, and that disagreement between the annotators is something that needs to be dealt with at all costs [@aroyo2015truth].
A recent study [@van2021validity] demonstrates that using crowd-coding platforms is a good way to collect such golden standards without too much disagreement.
These platforms have also been used for experiments, and the quality of the respondents has been a center of attention [@coppock2019generalizing; @clifford2015samples; @huff2015these; @berinsky2014separating].
While it has been argued that this data is of similar quality to data from a random sample of the population [@coppock2019generalizing], others have demonstrated that these online platforms are populated by people that are "unlike" the general public, being younger and holding more liberal values [@clifford2015samples; @huff2015these].
If the latter is true, this might cause a problem for the coding of political texts: @ennser2018impact report that coders of political texts incorporate prior beliefs about parties' issue stances into their coding decisions.
The authors find that party labels cue coders to a stance.
For example, coders are more likely to report a left-wing party to be pro-immigration and a populist right-wing party to be against based on the exact same sentence.
*Is this actually bias or a diversity of view points? And how big of a problem does this bias/diversity of view points generate for scholars relying on golden standard data?* This question also taps into newer developments related to data annotations: Increasingly, Large Language Models such as ChatGPT are being used by researchers [@gilardi2023chatgpt] as well as annotators themselves [@veselovsky2023artificial] to "simulate" human respondents for annotation tasks.
These models draw among other things from the gold standard data sets that have been created by researchers, perpetuating the specific biases that are present in them and not allowing for any further disagreement or diversity of viewpoints.

There is a long history of text annotation in studies analyzing political text.
While this yields lots of experiences as to how to train coders so that we get reliable hand-coded data, the procedure is expensive, protracted, and sometimes does not even get us the quality of data we need [@weber18].
While the crowd offers a solution to some of these issues [@van2021validity], the main underlying assumption that there is one correct interpretation for every input example remains untouched.
However, especially for more complex coding tasks different research fields have different interpretations of what counts as *correct* solution depending on the main goal of annotation: Do we want to create a clean benchmark data set that follows strict linguistic rules or one that reflects how statements are being interpreted by "real" readers that do not necessarily follow theoretical definitions on which elements are needed for a specified stance?
Trends of polarization have shown that people do interpret information according to their ideological position.
*Does this mean that some are right and others are wrong? Or is there an ideological difference in the ground truth?* These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced.
Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations [for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd, @ying2022topics].
However, work in computational linguistics has shown that increased annotation instructions do not increase quality [@parmar2022don].
This leaves us between a rock and a hard place.
Is there a potential bias in annotators that we should account for?

In this paper, we build upon the NLP literature on disagreement -- or bias -- in annotation [e.g., see @shen2021sounds; @geva2019we; @sommerauer2020penguin; @plank2014learning] and so-called perspectivism [@basile2021toward; @havens2022beyond] -- i.e. the adoption of methods that integrate the opinions and perspectives of the human subjects involved in the knowledge representation step of the machine learning processes [@basile2021toward].
This literature puts forward that disagreement can occur because of differences in ideological position or political knowledge [@shen2021sounds; @alkiek2022classification; @joseph2021mis].
This allows us to test the extent to which disagreement takes place, for what type of stances, as well as gives us directions on how to deal with the diversity in conceptions and the political heterogeneity that nowadays potentially occurs in our sample of annotators.
To do so, we have fielded two high-powered pre-registered experiments (see [here](link) and [here](link)) in the Netherlands -- a low-level polarized country -- and in the U.S. -- a high-level polarized country testing the effect of ideological distance between the annotator and the political actor in the text (H1), the effect of overinterpretation based on political knowledge or ideological engagement (H2), and an offered solution of masking the political actor to mitigate the effects of ideology and knowledge (H3).[^1]
In the experiments, we vary the level of specification with which a political actor takes a stance -- a declarative sentence versus a sentence where with some knowledge on politics, the stance might be inferred -- as well as whether the political actor is shown or masked with putting `[ACTOR]` instead of the political party.
We do this for four different political issues: *Environment*, *Immigration*, *Tax Policy*, and *EU* (for the Dutch case) or *Foreign Policy* (for the American case).
This country selection does not only allow us to showcase the scope conditions of disagreement due to different levels of political heterogeneity, but also differentiates between languages.
English is not only the most dominant language for computational text analysis in the social science [@baden2022three; Dolinsky et al. 2023], crowd-coders do not need to be native speakers, given the dominance of English in our daily lives.
This is different for Dutch, it is a language spoken by a smaller community, typically native speakers, yet still an often-enough researched case in computational text analysis in the social science [@baden2022three; Dolinsky et al. 2023].

@@

Our results demonstrate that overall sentences where with some knowledge on politics the stance might be inferred are really difficult for the crowd to annotate -- people overinterpret the position using their own knowledge of the world.
This is problematic as these sentences are very common in political text -- like legislative debates or speeches -- as well as media reports.
Moreover, our results also demonstrate that for these disagreements in the crowd to occur, the level of polarization needs to be high.
We do find support for our hypotheses in the American context, but not in the Dutch context -- except for the situation where masking overcomes political knowledge.
In this case, we do find support in the Dutch experiment, but not in the American one.
Our findings thus underline the importance of taking disagreement seriously for the creation of gold standard text -- the bread-and-butter of all machine learning endeavours.
We should look beyond the majority vote and model it in the data, because if an algorithm is trained on biased data from disagreeing annotators, it will reproduce and often exacerbate that bias when it is applied to new data [e.g., see @prost2019debiasing].
To be able to model these characteristic of annotators, we should survey the characteristics of annotators when using the crowd (see Webb-Williams et al. 2023 for a similar argument, yet different annotator characteristics).

@@

Generating large data-sets has become one of the main drivers of progress in natural language understanding.
In studies of political communication, the most familiar annotation tasks involve identifying basic concepts.
This includes noting the topic of the text, the position of the actor or the tone of the text.
A recent study [@van2021validity] demonstrates that using crowd-coding platforms is a good way to collect such large data-sets.
However, having only a few workers annotate the majority of text of interest has raised concerns about data diversity and models' ability to generalize beyond the crowd-workers: In a series of experiments, @geva2019we show that often models do not generalize well to annotations from annotators that did not contribute to the training set, suggesting that annotator bias should be monitored during data-set creation.
One such potential bias, especially in times of increasing polarization [@iyengar2019origins; @gidron2019toward; @boxell2022cross], is based on ideological position of the annotator.
Given that annotators on crowd-coding platforms tend to be younger and hold more liberal values than the general public [@clifford2015samples; @huff2015these], this could potentially hamper the data diversity and generalizability of the model.
An additional reason to monitor the annotators' ideological position as a potential source of annotator bias is that a recent study in NLP showed that experiential factors influence the consistency of how political ideologies are perceived [@shen2021sounds].
Their finding challenges the "ground-truth" assumption we as researcher make that a position for example is either left-leaning or right-wing leaning.
People with different ideological backgrounds might experience that position differently.
This challenges our way of data collection.
We are interested in the effect of e.g. elite communication.
To study this, we allow for heterogeneous treatment effects in experimental work.
This indicates that we often do not assume that the treatment, often using text, has the same effect for different partisans.
Yet, at the same time, we forget or ignore that knowledge when creating large data-sets for our machine learning models.

@@

Another aspect to consider is that disagreement can be informative for the concept under measure - sometimes agreement can be used to validate hypotheses about how universal the perceptions of such concepts are [@sommerauer2020penguin].
Additional doubts on smoothing out disagreement in annotation have focussed on the lack of diversity when only annotating with one label or annotator, leading to a homogeneity especially in subjective and social tasks [@geva2019we] such as hatespeech detection or political affiliation classification.
The question then is: *Whose perspective is being recorded in these datasets, and then later in the models trained on these datasets?* Framing arbitrary representations in data as "bias" misses the political character of datasets: There is no neutral data and no apolitical standpoint from where we can call out bias.
Datasets are always "a worldview" [26] and, as such, data always remains biased." (Miceli et al., 2022, p. 5). This is key to social scientist in general, and those studying political text in particular, since several tasks of interest are intrinsically societal, with answers that differ based on the make-up of the worldview of annotators. The answer of the annotators in turn influence how machine learning examples classify new models. For instance, hate speech detection and abuse detection is one NLP task where the race and gender of annotators influences annotators and model performance[@gordon2022, @Larimore, @waseem2016]. Language is inherently connected to society and culture: @shenDarling analyze sentiment analysis, and find that human annotators lead to certain perspectives on sentiment being recorded, and that notably that African American English dialects are often misunderstood by such models.

Most recently, new annotation paradigms have gone one step further by asking whether we are modelling the task, or the annotator [@geva2019we].
@pavlick2019 find that for the logical coherence task Natural Language Inference, annotators have several valid interpretations that are not reflected in one ground truth label.
They call for new training paradigms that can reflect "the full range of possible human inferences"[@pavlick2019, p. 688].
Recent approaches in NLP have sought to explicitly incorporate disagreement and diversity in training data annotations.
[@röttger2021] introduces the idea of an explicitly subjective annotation paradigm existing in addition to one focussed on one label and "ground truth".
Such a subjective annotation paradigm can be used for purposes where the goal is finding diverse perspectives on the task or concepts, and for models to model more accurately how humans interpret a task or text.
Additionally, "perspectivism" [@basile2021toward] is a paradigm and research agenda where different perspectives are explicitly incorporated in the training data, and used by models to provide more human-like classifications.
Another paradigm is "jury learning" [@gordon2022], in which machine learning models do not learn to replicate one specific ground truth, but are trained with different annotator juries to reflect the judgement of different populations.
In both approaches, demographic and other individual aspects of the annotator are explicitly mentioned and highlighted as having an influence on classification performance, but is used as an asset rather than as an error.

@troiano2021 [p.2] also note how for a complex concept such as the "emotion" of a text, the annotator can make several assumptions during the annotation process on what is wanted, that are all valid and may or may not be useful in different contexts: "It is possible to assess one's own emotion after reading the text, to reconstruct the affective state of the writers who produced it, to guess the reaction that they intended to elicit in the readers, and so on." @shen2021sounds find that political ideology is not an inherent concept in many texts, but rather dependent on who is asked to annotate and their perceptions and background.
Extralinguistic factors, such as annotator's own political ideology and also knowledge, influenced annotation and in turn model performance.
@thornjakobsen2022 specifically analyze a task related to stance detection, argumentative sentence detection, on these extralinguistic factors, and find an effect of gender and political leaning on annotations and also model performance.
However, to our knowledge this phenomenon has not yet been tested in a controlled experimental setting with manipulations in the texts.

@@

> **Ideological Bias hypothesis** (*H1b*): The effect of H1a is stronger for sentences in which the party's position can be inferred (i.e. underspecified sentences).

As noted by @plank2014, some linguistic examples are ambiguous, and open to multiple interpretations even for seemingly objective tasks such as part of speech tagging, where annotators have to distinguish parts of speech such as nouns and verbs.
A complex concept such as political ideology is much more likely to lead to multiple interpretations.
We call such sentences with more possible interpretations and less explicit standpoints "underspecified".
A lack of explicitness in the annotated text is one of the main causes of the disagreements in earlier literature.
@thornjakobsen2022 deduce that annotator bias comes from a process known as the affect heuristic [@slovic2007affect ]: making a decision based on the emotional response related to your own personal attitude towards the discussed topic, especially when the text is relatively ambiguous.
In the tradition of more strict interpretations of what constitutes as a stance especially in (computational) linguistics, sentences that are underspecified should always be annotated as "not a stance".
However, this might not be the desired annotation for other research purposes: When for example the main goal is to understand how readers are affected by statements shown in e.g., a newspaper article, a more lenient definition of stance that allows annotators to infer the direction of a political stance from context and prior knowledge even though a statement is strictly speaking underspecified might be more useful.
In everyday life, people will not follow strict linguistic definitions, for understanding media effects we might thus be more interested in whether stances, giving context information, are "correctly overinterpreted".
To test whether people with different ideological backgrounds as well as their political knowledge might experience underidentified position differently, challenging our ground-truth assumption in data annotation, we propose the following hypotheses:

@@

We will test these hypotheses both with a strict and a lenient interpretation of stance to illustrate whether decisions made in the research process on what constitutes a stance influence the results.
In a next step, we ask: *If these biases exists, how can we alleviate them?* There have been several previous approaches to solve biased annotations, especially where it concerns political or societal aspects.
@geva2019we introduce an approach where training set annotators are separated from annotators annotating data sets that evaluate the models, to ensure the evaluation is not simply accurate at replicating the original annotators, but can generalize to new annotators' judgements.
However, these approaches are not aimed at reducing biases during the training set annotation procedure.
Other approaches are aimed at leveraging multiple perspectives - but this is not useful when one wants one label to learn from.
For the Austrian National Elections, @ennser2018impact already demonstrated that showing party labels impact annotators' assessment of the party position.
So, one solution is masking:

@@

We have conducted the survey experiments in the Netherlands in May 2022 and in the United States in January 2023.
Both samples, recruited through [KiesKompas](https://www.kieskompas.nl/en/) and [Prolific](https://www.prolific.co/) for respectively the Dutch and American case, consist of 3,000 participants (based on the power analysis presented in our [online compendium](LINK)) of 18 years and older.
Both survey companies works with non-random opt-in respondents.
Therefore, we measured many demographic background variables, and balance checks have been conducted to demonstrate whether certain categories are over represented in a certain experimental group.
Our study has been approved by the [Research Ethics Review Committee](https://fsw.vu.nl/nl/onderzoek/research-ethics-review/index.aspx) of the *Vrije Universiteit Amsterdam* (see the approval [here](https://github.com/MarikenvdVelden/bias-experiment/blob/master/docs/2022-3-30-59.pdf)).
To ensure good quality of our data, one attention check (discussed in more detail in [Section 3.3](#attention-checks)) is included [@berinsky2014separating].

@@

***Moderating Covariates.*** *Ideological position* is measured using an 11-point scale ranging from left (`0`) to right (`10`).
*Political knowledge* is measured with six items from the Dutch Parliamentary Election Studies for the Dutch sample, and the three items from the American National Election Studies.[\^]:The questionnaire can be found in OA **XX**.

***Control Variables.*** In our analysis, we control for demographic information (gender, age, education, income, religion, job) as well as political background variables (trust in politics, ideological position on economic left-right scale and cultural progressive-conservative scale, and evaluations and prospects of the economy).
Tables A.5 till A.17 in the OA demonstrate the descriptive information per country.
